Vibronic Coupling Effects in the Photoelectron Spectrum of Ozone: A Coupled-Cluster Approach

### 2024-09-02 article.tex
Update Resuts and Disucussion with Dr. Stanton comments

\documentclass[
12pt,
a4paper,
prb,
superscriptaddress,
tightenlines,  % single-spaced
]{revtex4}
\usepackage{amsmath,amssymb,physics}
\usepackage{graphicx} % For figures with images
\usepackage{natbib} % bibliography
% \setcitestyle{super, sort&compress, comma} % chemistry style (prb's default)
\usepackage{hyperref} % Clickable links in the document and their colors
\hypersetup{
    colorlinks=true, % false: boxed links; true: colored links
    linkcolor=cyan, % color of internal links
    citecolor=magenta, % color of links to bibliography
    filecolor=magenta, % color of file links
    urlcolor=cyan, % color of external links
    runcolor=cyan
}

@@

\begin{abstract}
One of the most important areas of application for equation-of-motion coupled
cluster (EOM-CC) theory is to the prediction, simulation and analysis of
various types of electronic spectra.   In this work, the EOM-CC method known
as EOMIP-CC is applied to the closely lying and coupled pair of states of the
ozone cation -- ${\tilde X}^2A_1$ and ${\tilde A}^2B_2$ -- using sophisticated
approaches that extend to the full singles, doubles, triples and quadruples
model (EOMIP-CCSDTQ).   Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOMIP-CC
calculations, this research provides a spectrum that is in good agreement with
the photoelectron spectrum of ozone.   An important result here is that the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be denied that water
(H$_2$O) is the most important, the most highly studied, and the most
well-understood.  Beyond H$_2$O, there are many triatomic molecules that have
an environmental, technological or biological importance, have been subjected
to many studied and are understood to various levels of detail.  Perhaps the
most interesting such case is ozone (O$_3$), which has a vast number of
important properties, a very rich history of study \cite{chappuis}, and –
unlike the relatively simple water molecule – a profound quantum-mechanical
complexity \cite{Babikov:anomalousOzone:2003}.  In the latter context, while
we think of (and an NMR experiment would reveal) as two distinct kinds of
oxygen atoms in ozone, the full molecular Hamiltonian does not distinguish
between them types, with the three equivalent structures separated by a
barrier that lies tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O
dissociation threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}. In reality, the
energy levels of ozone all have a near triple ($e+a$) degeneracy, albeit with
a tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}).   More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

@@

Our contribution to this issue paying homage to the career and accomplishments
of R.J. Bartlett consists of an application of a vibronic coupling model,
parametrized by EOM-CC calculations, to the photoelectron spectrum of ozone.
We trust that this combination of methodology applied to a molecule that has
been extensively studied by the Bartlett group is an appropriate contribution
to this issue.

\begin{figure}
    \includegraphics[width = 16 cm]{./figures/ozone_intro}
    \caption{ 
        Two lowest states of the ozone cation and normal modes of ozone.
    }
    \label{fig:ozone_intro}
\end{figure}

@@

Ozone is a $C_{2v}$ molecule (we are following the Mulliken's
convention~\cite{Mulliken:55:symnot} by placing the molecule in the $yz$ plane
and aligning the molecular symmetry axis with the $z$ direction) with three
normal modes: symmetric stretch, symmetric bend, and asymmetric stretch.  The
asymmetric stretch is of a $b_2$ symmetry. The two lowest electronic states of
the ozone cation are close in energy. The lower state is fully symmetric while
the higher one is of the $B_2$ symmetry, the same symmetry as the asymmetric
stretch. The close energetic separation and the matching symmetry of the
asymmetric stretch results in significant vibronic effects appearing in the
ozone photoelectron spectrum.  Figure~\ref{fig:ozone_intro} presents a
graphical summary of this information.

We simulate the vibronic states of the ozone cation using the model
Hamiltonian of K{\"o}ppel, Domcke, and Cederbaum (KDC
Hamiltonian).\cite{Cederbaum:LVC:84,KDC:81,Koppel:CIbookCh7:04} This is a
multi-state and multi-mode Hamiltonian defined in the basis of diabatic
states.  We consider the ozone cation in the basis of two quasi-diabatic
states coupled by one mode (mode number 3). The model also includes two
symmetric modes (modes number 1 and 2)
\begin{subequations}
    \begin{equation}
        H = H _0 \mathbf{1}
        +
        \begin{pmatrix}
            V ^{(1)}  & \lambda _3 Q _3\\
            \lambda _3 Q _3 & V ^{(2)}
        \end{pmatrix}
    \end{equation}
    \begin{equation}
        H _0 = 
        \frac{1}{2} \left(\sum _{i = 1}^3 
        - \omega _i \frac{\partial ^2}{\partial Q _i ^2 }\right)
        + \frac{1}{2}\omega _3 Q _3 ^2
    \end{equation}
    \begin{equation}
        V ^{(\alpha)} = 
        E ^{(\alpha)} 
        + \sum _{i,j,k,l \in \{1,2\}} 
            \kappa ^{(\alpha)} _i Q _i 
            + \kappa ^{(\alpha)} _{ij} Q _i Q _j 
            + \kappa ^{(\alpha)} _{ijk} Q _i Q _j Q _k 
            + \kappa ^{(\alpha)} _{ijkl} Q_i Q _j Q _k Q _l.
    \end{equation}
    \label{eq:KDC}
\end{subequations}
The Hamiltonian parameters are expanded around the equilibrium geometry of
ozone. $E ^{(\alpha)}$ are the vertical ionization energies calculated at that
geometry. $Q_i$ are the dimensionless normal coordinates of ozone. $\kappa$ are
the coefficients of expansion of the potential along the fully symmetric
coordinates. $\lambda$ are the linear diabatic couplings.  $\omega$ are the
harmonic frequencies of ozone.

We find the parameters that enter the KDC Hamiltonian using \emph{ab inito}
coupled-cluster (CC) methods and its equation-of-motion EOM-CC
extension.~\cite{Bartlett:CC_review:07, Krylov:EOMRev:07, Bartlett:Book:09,
Christiansen:EOMRev:11, Bartlet:EOMRev:12, Krylov:OSRev} We use the CC
truncated to singles and doubles (CCSD), singles, doubles and triples (CCSDT)
as well as singles, doubles, triples and quadruples
(CCSDTQ).\cite{Matthews:ncc:2015}  We use the EOM-CC for ionization potential
(EOM-IP).~\cite{StantonGauss:EOMIP:99} We use the Ichino, Gauss and Stanton
definition of quasi-diabatic states based on the EOM-CC method
(EOM-CC-QD)~\cite{Stanton:EOMIPdeg:09}. In all CC and EOM-CC calculations we
leave all electrons correlated, in other words, we do not use the
frozen-core approximation.

Using CCSDT/ANO1 we optimize the geometry of ozone and compute its harmonic
frequencies and normal coordinates.~\cite{Almlof:ANO,Almlof:ANO:1988} At the
same geometry we compute the linear diabatic coupling $\lambda$ using
EOM-IP-CCSD-QD/ANO1.  We use EOM-IP-CCSDT/ANO1 on a grid to find the
expansion coefficients $\kappa$.

The vertical ionization energies are calculated using a composite method. The
base value is the complete basis set (CBS) extrapolation of the
EOM-IP-CCSDT/cc-pCVnZ energies with n = 5, 6.~\cite{Woon:95:CCBS} These values
are augmented with two corrections: the $\Delta$Q correction in the cc-pwCVTZ
basis set and the relativistic correction calculated using
EOM-IP-CCSD/cc-pwCVTZ.~\cite{Dunning:02:p(w)CVnZ} We introduce an error
estimates to the reported vertical energies. For the error estimate of the
extrapolated CBS value we use half of the absolute value of the difference
between the best \emph{ab initio} value and the extrapolated value. For the
remaining corrections we use as the error estimate half of the absolute value
of the correction.

To compare the simulated photoelectron spectrum to the experimental one, we
use the oscillator strengths ratio A$_1$:B$_2$ of $1$:$1.35$.\cite{KDC:O3:92}
Additionally, the stick spectrum is broadened with the Lorenzian envelopes
normalized to the peaks' intensities
\begin{equation}
    f _{envelope}(x, x _i, I _i) = 
    \frac{I_i}{(x-x _i)^2 + (\gamma/ 2) ^2}.
    \label{eq:lorentzian}
\end{equation}
$x _i$ is the position of the spectral peak, $I _i$ is its intensity and
$\gamma$ is the peak's width.

We use the \textsc{xsim} program of Dr. Stanton to simulate the spectrum using
the basis of 50 harmonic states in each mode and 6000 iterations of the
Lanczos procedure. All remaining \emph{ab initio} calculations are completed
using \textsc{CFOUR}.~\cite{cfour, cfour:2020}

@@

The optimized geometry of ozone gives the bond lengths of $1.270$~\AA{} and
the bond angle of $116.9^\circ$. The two symmetric normal modes have
frequencies $1169$~cm$^{-1}$ and $724$~cm$^{-1}$, while the asymmetric stretch
has frequency equal to $1129$~cm$^{-1}$. The computed value of the linear
diabatic coupling constant $\lambda$ is $1394$~cm$^{-1}$. The vertical
ionization energy for the first excited state, $E^{(1)}$, computed using the
composite method described above is equal to $12.827$~eV. Our error estimate
for this value is $30$~meV, see Table~\ref{tab:vertical_ionization_energy} for
details.  We note that the convergence of the vertical ionization gap between
the two states is much faster and we estimate this value as equal to
$123\pm8$~meV, see Table~\ref{tab:vertical_gap} for details.

@@

\begin{table}
    \caption{
        The energy gap between vertical ionization energies of the $^2$A$_1$
        and $^2$B$_2$ states with error estimates, meV. See caption of
        Table~\ref{tab:vertical_ionization_energy}.
    }
    \label{tab:vertical_gap}
    \begin{center}
        \begin{tabular}[c]{|l|r|r|}
            \hline
            Contribution             &  Gap    & Error estimate \\ \hline
            EOM-CCSDT/CBS            &  108.8  & 0.9 \\
            $\Delta$Q/pwCVTZ         &   16.5  & 8 \\
            Relativistic/CCSD/pwCVTZ &   -2.2  & 1.1 \\ \hline
            Final value, meV         &  123    & 8 \\
            Final value, cm$^{-1}$   &  990    & 65 \\ \hline
        \end{tabular}
    \end{center}
\end{table}

@@

\begin{figure}
    \includegraphics[width=10 cm]{figures/spectrum_overline}
    \caption{
        Simulated photoelectron spectrum of ozone. Bottom axis shows energy
        scale in eV. Top axis shows energy offset from the origin in
        cm$^{-1}$. Stick spectrum shows positions and intensities of all
        simulated states. Blue color corresponds to the oscillator strength
        originating from the $^2$A$_1$ basis state while the red one
        corresponds to $^2$B$_2$. Gray vertical lines with captions on top
        indicate positions of features as measured by the PFI-ZEKE
        experiment.~\cite{Willitsch:O3ZEKE:2005} The simulated spectrum was
        shifted to match the PFI-ZEKE experimental origin. $D_0$ marks the
        dissociation threshold of O$_3^+$. Gray dotted line marks the energy
        of the minimum of the conical intersection (CI).
    }
    \label{fig:ozone_overlay}
\end{figure}

\begin{figure}
    \includegraphics[width=10cm]{figures/spectrum_assigned.pdf}
    \caption{
        Simulated photoelectron spectrum of ozone, but with no account given
        to the vibronic coupling effect.
    }
    \label{fig:no_coupling}
\end{figure}

We would like to assign the vibronic peaks from our simulation. To this end,
we run the simulation once again, this time, however we set the linear
diabatic constant to zero, i.e., $\lambda = 0$. With that change we can
reproduce the spectrum (at an equivalent level of theory) but in an artificial
case where there is no vibronic coupling. Figure~\ref{fig:no_coupling}
presents this spectrum. The non-coupled spectrum is easy to assign using the
labels that mark the symmetry of the electronic state, A$_1$ or B$_2$, and the
vibrational state ($\nu _1 \nu_2 \nu_3$), where $\nu _i$ is the number of
quanta in mode $i$ with $i=1$ for the symmetric stretch, $i=2$ for the
symmetric bend and $i=3$ for the asymmetric stretch. The assigned spectrum
shows progressions in the symmetric bend. There is one such progression in
each electronic state.  Additionally for each state, there is also another
progression in the symmetric bend with one excitation in the symmetric
stretch.

@@

We decompose the vibronic states of our main simulation in the basis of the
artificial, uncoupled states discussed in the previous paragraph.
Figure~\ref{fig:sim_vs_zeke} shows the assigned spectrum and
Table~\ref{tab:peak_assignment} lists the decomposition of all peaks in the
region of up to $3000$~cm$^{-1}$ with intensities larger than $10^{-3}$. We
compare the assigned spectrum against the high-resolution
pulsed-field-ionization zero-kinetic-energy (PFI-ZEKE) spectrum from
2005.\cite{Willitsch:O3ZEKE:2005} The PFI-ZEKE spectrum is the best source of
information on the location of the peaks, especially the origin, against which
align our simulation. The origin of the simulated spectrum is located
$170$~cm$^{-1}$ lower than the experimental origin which was experimentally
observed at $101,020.5$~cm$^{-1}$.\cite{Willitsch:O3ZEKE:2005} This is within
our error estimate for the vertical ionization energy
($30$~meV$=240$~cm$^{-1}$).

@@

Lastly, we compare our simulation to an earlier accurate simulation, the work
of Tarroni and Carter from 2011.~\cite{tarroni:O3:2011}
Figure~\ref{fig:sim_vs_tarronicarter} presents a comparison of lines from both
simulations. The comparison includes only the lines from the earlier
simulation that were tabulated with an assignment.

\begin{table}
    \renewcommand{\arraystretch}{0.8}% Tighter
    \caption{Assignment and decomposition of the eigenvectors of the ozone
    cation.}
    \label{tab:peak_assignment}
    \begin{tabular}{|l|l|l|}
        \hline
        Peak position (cm$^{-1}$) & Assignment & Eigenvector \\
        \hline
$     0$ &  $A_1(000)$ & $ -0.99~A_1(000)$ \\
$   618$ &  $A_1(010)$ & $ +0.99~A_1(010)$ \\
$   915$ &  $A_1(001)$ & \\
$  1076$ &  $A_1(100)$ & $ -0.99~A_1(100)$ \\
$  1222$ &  $A_1(020)$ & $ -0.97~A_1(020)$ \\
$  1368$ &  $B_2(000)$ & $ +0.97~B_2(000)$ \\
$  1498$ &             & $ -0.10~B_2(000)$ \\
$  1684$ &  $A_1(110)$ & $ -0.96~A_1(110)$ \\
$  1796$ &  $A_1(030)$ & $ -0.87~A_1(030) -0.16~A_1(110) -0.13~A_1(020) +0.13~A_1(040)$ \\
$  1848$ &             & $ +0.31~A_1(030)$ \\
$  1921$ &  $B_2(010)$ & $ -0.74~B_2(010) +0.18~B_2(020) +0.11~B_2(000)$ \\
$  1977$ &             & $ +0.24~B_2(010)$ \\
$  2085$ &             & $ -0.52~B_2(010)$ \\
$  2132$ &             & $ +0.25~A_1(030) +0.24~A_1(040) +0.13~A_1(120) -0.13~A_1(050)$ \\
$  2275$ &  $A_1(120)$ & $ +0.86~A_1(120) +0.12~A_1(040)$ \\
$  2363$ &             & $ -0.62~A_1(040) +0.40~A_1(120) -0.17~A_1(030) +0.11~A_1(050)$ \\
$  2439$ &             & $ +0.47~A_1(040) +0.12~A_1(120)$ \\
$  2454$ &             & $ +0.53~B_2(020) +0.24~B_2(010) -0.17~B_2(030)$ \\
$  2576$ &             & $ +0.27~B_2(020)$ \\
$  2736$ &             & $ +0.60~B_2(020)$ \\
$  2886$ &  $A_1(130)$ & $ -0.81~A_1(130) +0.12~A_1(050) -0.10~A_1(120)$ \\
$  2974$ &             & $ -0.24~B_2(110) -0.16~B_2(020)$ \\
$  3047$ &  $A_1(050)$ & $ +0.76~A_1(050) +0.13~A_1(130)$ \\
        \hline
    \end{tabular}
\end{table}

@@

We first discuss the comparison of our simulations to the lower-resolution
experimental spectrum presented on Figure~\ref{fig:sim_vs_dyke}. The peak A is
known to be a hot band.~\cite{KDC:O3:92} The simulation reproduces well the
consecutive increase in the intensities of peaks B, C, D, and E. The spacing
between these peaks is also well reproduced. Drop in the intensity at the peak
F is captured by the simulation. Starting from the peak G, the simulation
shows discrepancy with the experiment. A sudden drop in the intensity past the
peak H is not observed in the simulation. We discuss a likely source of this
mismatch later.

>>>>

\documentclass[
12pt,
a4paper,
prb,
superscriptaddress,
% tightenlines,  % single-spaced
]{revtex4}
\usepackage{amsmath,amssymb,physics}
\usepackage{graphicx} % For figures with images
\usepackage{natbib} % bibliography
% \setcitestyle{super, sort&compress, comma} % chemistry style (prb's default)
\usepackage{hyperref} % Clickable links in the document and their colors
\hypersetup{
    colorlinks=true, % false: boxed links; true: colored links
    linkcolor=cyan, % color of internal links
    citecolor=magenta, % color of links to bibliography
    filecolor=magenta, % color of file links
    urlcolor=cyan, % color of external links
    runcolor=cyan
}

@@

\begin{abstract}
One of the most important areas of application for equation-of-motion coupled
cluster (EOM-CC) theory is the prediction, simulation and analysis of
various types of electronic spectra.   In this work, the EOM-CC method known
as EOMIP-CC is applied to the closely lying and coupled pair of states of the
ozone cation -- ${\tilde X}^2A_1$ and ${\tilde A}^2B_2$ -- using sophisticated
approaches that extend to the full singles, doubles, triples and quadruples
model (EOMIP-CCSDTQ).   Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOMIP-CC
calculations, this research provides a spectrum that is in good agreement with
the photoelectron spectrum of ozone.   An important result here is that the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be denied that water
(H$_2$O) is the most important, the most highly studied, and the most
well-understood.  Beyond H$_2$O, there are many triatomic molecules that have
an environmental, technological or biological importance, have been subjected
to many studied and are understood to various levels of detail.  Perhaps the
most interesting such case is ozone (O$_3$), which has a vast number of
important properties, a very rich history of study \cite{chappuis}, and 
unlike the relatively simple water molecule, a profound quantum-mechanical
complexity \cite{Babikov:anomalousOzone:2003}.  In the latter context, while
we think of (and an NMR experiment would reveal) as two distinct kinds of
oxygen atoms in ozone, the full molecular Hamiltonian does not distinguish
between them types, with the three equivalent structures separated by a
barrier that lies tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O
dissociation threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}. In reality, the
energy levels of ozone all have a near triple ($e+a$) degeneracy, albeit with
a tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}).   More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

@@

Our contribution to this issue paying homage to the career and accomplishments
of R.J. Bartlett consists of an application of a vibronic coupling model,
parametrized by EOM-CC calculations, to the photoelectron spectrum of ozone.
We trust that this combination of methodology applied to a molecule that has
been extensively studied by the Bartlett group is an appropriate contribution.

\begin{figure}
    \includegraphics[width = 16 cm]{./figures/ozone_intro}
    \caption{ 
        Two lowest states of the ozone cation and normal modes of the ozone
        neutral.
    }
    \label{fig:ozone_intro}
\end{figure}

@@

Ozone is a $C_{2v}$ molecule (following Mulliken's
convention~\cite{Mulliken:55:symnot} the molecule is placed in the $yz$-plane
and the molecular symmetry axis aligns with the $z$ direction) with three
normal modes: symmetric stretch, symmetric bend, and asymmetric stretch.  The
asymmetric stretch is of $b_2$ symmetry. The two lowest electronic states of
the ozone cation are close in energy. The lower state is fully symmetric while
the higher one is of $B_2$ symmetry, the same symmetry as the asymmetric
stretch. The close energetic separation and the matching symmetry of the
asymmetric stretch is accompanied by significant vibronic effects in the
lowest two states of the ozone photoelectron.  Figure~\ref{fig:ozone_intro}
presents a graphical summary of this information.

The vibronic states of the ozone cation are simulated using the model
Hamiltonian of K{\"o}ppel, Domcke, and Cederbaum (KDC
Hamiltonian).\cite{Cederbaum:LVC:84,KDC:81,Koppel:CIbookCh7:04} This is a
multi-state and multi-mode Hamiltonian defined in the basis of diabatic
states. The ozone cation is considered in the basis of two quasi-diabatic
states coupled by one mode (mode number 3). The model also includes two
symmetric modes (modes number 1 and 2)
\begin{subequations}
    \begin{equation}
        H = H _0 \mathbf{1}
        +
        \begin{pmatrix}
            V ^{(1)}  & \lambda _3 Q _3\\
            \lambda _3 Q _3 & V ^{(2)}
        \end{pmatrix}
    \end{equation}
    \begin{equation}
        H _0 = 
        \frac{1}{2} \left(\sum _{i = 1}^3 
        - \omega _i \frac{\partial ^2}{\partial Q _i ^2 }\right)
        + \frac{1}{2}\omega _3 Q _3 ^2
    \end{equation}
    \begin{equation}
        V ^{(\alpha)} = 
        E ^{(\alpha)} 
        + \sum _{i \in \{1,2\}} 
            \kappa ^{(\alpha)} _i Q _i 
        + \sum _{i,j \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ij} Q _i Q _j 
        + \sum _{i,j,k \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijk} Q _i Q _j Q _k 
        + \sum _{i,j,k,l \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijkl} Q_i Q _j Q _k Q _l.
    \end{equation}
    \label{eq:KDC}
\end{subequations}
The vibronic Hamiltonian elements are expanded around the equilibrium geometry
of neutral ozone. $E ^{(\alpha)}$ are the vertical ionization energies
calculated at that geometry. $Q_i$ are the dimensionless normal coordinates of
ozone, $\kappa$ are the coefficients of expansion of the potential along the
fully symmetric coordinates, and $\lambda$ the linear diabatic coupling,
$\omega$ represent the harmonic frequencies of the neutral.

The parameters that enter the KDC Hamiltonian are computed using \emph{ab
inito} coupled-cluster (CC) methods and its equation-of-motion EOM-CC
extension.~\cite{Bartlett:CC_review:07, Krylov:EOMRev:07, Bartlett:Book:09,
Christiansen:EOMRev:11, Bartlet:EOMRev:12, Krylov:OSRev} The CC expansion is
truncated to singles and doubles (CCSD), singles, doubles and triples (CCSDT)
as well as singles, doubles, triples and quadruples
(CCSDTQ).\cite{Matthews:ncc:2015}  The states of the cation are described with
EOM-CC for ionization potential (EOMIP).~\cite{StantonGauss:EOMIP:99} The
definition of quasi-diabatic states of Ichino, Gauss and Stanton based on the
EOMIP-CC method (EOMIP-CC-QD) is used here~\cite{Stanton:EOMIPdeg:09}. All CC
and EOM-CC calculations correlate all electrons, in other words, the
frozen-core approximation is not invoked.

The geometry of the neutral ozone and the computation of its harmonic
frequencies and normal coordinates uses
CCSDT/ANO1.~\cite{Almlof:ANO,Almlof:ANO:1988} 
The linear diabatic coupling $\lambda$ is computed at the same geometry using
EOMIP-CCSD-QD/ANO1. The expansion coefficients $\kappa$ are computed using the
EOMIP-CCSD/ANO1 energies calculated on a grid around the same geometry.

The vertical ionization energies are calculated using a composite method.  The
starting value is the complete basis set (CBS) extrapolation of the
EOM-IP-CCSDT/cc-pCVnZ energies with n = 5, 6.~\cite{Woon:95:CCBS} These values
are augmented with two corrections: the $\Delta$Q correction in the cc-pwCVTZ
basis set and the relativistic correction calculated using
EOM-IP-CCSD/cc-pwCVTZ.~\cite{Dunning:02:p(w)CVnZ} The error estimate of the
extrapolated CBS reference values is reported using half of the absolute value
of the difference between the best \emph{ab initio} value and the extrapolated
value.  For the remaining corrections (quadruples correction, $\Delta Q$, and
the relativistic correction) the error estimate is reported as half of the
absolute value of the correction.

To compare the simulated photoelectron spectrum to experiment, the
cross-section ratio A$_1$:B$_2$ of $1$:$1.35$ is used.\cite{KDC:O3:92}
Additionally, the stick spectrum is broadened with the Lorenzian envelopes
normalized to the feature intensities
\begin{equation}
    f _{envelope}(x, x _i, I _i) = 
    \frac{I_i}{(x-x _i)^2 + (\gamma/ 2) ^2}.
    \label{eq:lorentzian}
\end{equation}
$x _i$ is the position of the spectral peak, $I _i$ is its intensity and
$\gamma$ is the peak's width.

The spectrum is simulated using the \textsc{xsim} program in \textsc{CFOUR},
with the basis of 50 harmonic states in each mode and 6000 iterations of the
Lanczos procedure.~\cite{Sharma:xsim_socjt:2024} All calculations are
completed using \textsc{CFOUR}.~\cite{cfour, cfour:2020}

@@

The optimized geometry of ozone gives a bond length of $1.270$~\AA{} and a
bond angle of $116.9^\circ$. The two symmetric normal modes have frequencies
$1169$~cm$^{-1}$ and $724$~cm$^{-1}$, while the asymmetric stretch has a
frequency equal to $1129$~cm$^{-1}$. The computed value of the linear diabatic
coupling constant $\lambda$ is $1394$~cm$^{-1}$. The vertical ionization
energy for the first excited state, $E^{(1)}$, computed using the composite
method described above is  $12.827$~eV. Our error estimate for this value is
$30$~meV, see Table~\ref{tab:vertical_ionization_energy} for details.  We note
that the convergence of the vertical gap between the two states is much faster
and this value is estimated as $123\pm8$~meV, see Table~\ref{tab:vertical_gap}
for details.

@@

\begin{table}
    \caption{
        The energy gap between vertical ionization energies of the $^2$A$_1$
        and $^2$B$_2$ states with error estimates, meV. See caption of
        Table~\ref{tab:vertical_ionization_energy}.
    }
    \label{tab:vertical_gap}
    \begin{center}
        \begin{tabular}[c]{|l|r|r|}
            \hline
            Contribution             &  Vertical gap    & Error estimate \\ \hline
            EOM-CCSDT/CBS            &  108.8           & 0.9 \\
            $\Delta$Q/pwCVTZ         &   16.5           & 8 \\
            Relativistic/CCSD/pwCVTZ &   -2.2           & 1.1 \\ \hline
            Final value, meV         &  123             & 8 \\
            Final value, cm$^{-1}$   &  990             & 65 \\ \hline
        \end{tabular}
    \end{center}
\end{table}

@@

\begin{figure}
    \includegraphics[width=10 cm]{figures/spectrum_overline}
    \caption{
        Simulated photoelectron spectrum of ozone. Bottom axis shows energy
        scale in eV. Top axis shows energy offset from the origin in
        cm$^{-1}$. Stick spectrum shows positions and intensities of all
        simulated states. Colors blue and orange mark states of A$_1$ and
        B$_2$ symmetry, respectively. Gray vertical lines with captions on top
        indicate positions of features as measured by the PFI-ZEKE
        experiment.~\cite{Willitsch:O3ZEKE:2005} The simulated spectrum was
        shifted to match the PFI-ZEKE experimental origin. $D_0$ marks the
        dissociation threshold of O$_3^+$.~\cite{Willitsch:O3ZEKE:2005}
        Gray dotted line marks the energy of the minimum of the conical
        intersection (CI) located by our Hamiltonian.
    }
    \label{fig:ozone_overlay}
\end{figure}

\begin{figure}
    \includegraphics[width=10cm]{figures/spectrum_assigned.pdf}
    \caption{
        Simulated Franck-Condon photoelectron spectrum of ozone, vibronic
        coupling removed.
    }
    \label{fig:no_coupling}
\end{figure}

The vibronic peaks are assigned using a comparison to the Franc-Condon
simulation.  To this end, the spectrum is simulated once again, this time,
however, setting the linear diabatic constant to zero, i.e., $\lambda = 0$.
That change reproduces the spectrum (at an equivalent level of theory) but in
an artificial case where there is no vibronic coupling, i.e., the
Franck-Condon spectrum.  Figure~\ref{fig:no_coupling} presents this spectrum.
The non-coupled spectrum is easy to assign using the labels that mark the
symmetry of the electronic state, A$_1$ or B$_2$, and the vibrational state
($\nu _1 \nu_2 \nu_3$), where $\nu _i$ is the number of quanta in mode $i$
with $i=1$ for the symmetric stretch, $i=2$ for the symmetric bend and $i=3$
for the asymmetric stretch.  The assigned spectrum shows progressions in the
symmetric bend. There is one such progression in each electronic state.
Additionally for each state, there is also another progression in the
symmetric bend with one excitation in the symmetric stretch.

@@

The vibronic states are decomposed in the basis of the artificial, uncoupled
states discussed in the previous paragraph.  Figure~\ref{fig:sim_vs_zeke}
shows the assigned spectrum and Table~\ref{tab:peak_assignment} lists the
decomposition of all peaks in the region of up to $3000$~cm$^{-1}$ with
intensities larger than $1/300$ of the maximum peak. 
The resulting assignment is compared against the high-resolution
pulsed-field-ionization zero-kinetic-energy (PFI-ZEKE) spectrum from
Ref.~\cite{Willitsch:O3ZEKE:2005}. The PFI-ZEKE spectrum is the best source of
information on the location of the peaks, especially the origin, against which
the origin peak of the simulation is aligned. The origin of the simulated
spectrum is located $170$~cm$^{-1}$ lower than the experimental origin which
was seen at $101 020.5$~cm$^{-1}$.\cite{Willitsch:O3ZEKE:2005} This is within
our error estimate for the vertical ionization energy
($30$~meV$=240$~cm$^{-1}$).

@@

Lastly, a comparison to an earlier accurate simulation, the work of Tarroni
and Carter, Ref.~\cite{Willitsch:O3ZEKE:2005}, is presented on
Figure~\ref{fig:sim_vs_tarronicarter}. The comparison includes only the lines
from the earlier simulation that were tabulated with an assignment.

\begin{table}
    \renewcommand{\arraystretch}{0.8}% Tighter
    \caption{Assignment and decomposition of the eigenvectors of the ozone
    cation.}
    \label{tab:peak_assignment}
    \begin{tabular}{|l|l|l|l|}
        \hline
        Energy (cm$^{-1}$) & Symmetry& Assignment & Eigenvector \\
        \hline
$     0$ & $A_1$ &  $A_1(000)$ & $ -0.99~A_1(000)$ \\
$   618$ & $A_1$ &  $A_1(010)$ & $ +0.99~A_1(010)$ \\
$   915$ & $B_2$ &  $A_1(001)$ & \\
$  1076$ & $A_1$ &  $A_1(100)$ & $ -0.99~A_1(100)$ \\
$  1222$ & $A_1$ &  $A_1(020)$ & $ -0.97~A_1(020)$ \\
$  1368$ & $B_2$ &  $B_2(000)$ & $ +0.97~B_2(000)$ \\
$  1498$ & $B_2$ &             & $ -0.10~B_2(000)$ \\
$  1684$ & $A_1$ &  $A_1(110)$ & $ -0.96~A_1(110)$ \\
$  1796$ & $A_1$ &  $A_1(030)$ & $ -0.87~A_1(030) -0.16~A_1(110) -0.13~A_1(020) +0.13~A_1(040)$ \\
$  1848$ & $A_1$ &             & $ +0.31~A_1(030)$ \\
$  1921$ & $B_2$ &  $B_2(010)$ & $ -0.74~B_2(010) +0.18~B_2(020) +0.11~B_2(000)$ \\
$  1977$ & $B_2$ &             & $ +0.24~B_2(010)$ \\
$  2085$ & $B_2$ &             & $ -0.52~B_2(010)$ \\
$  2132$ & $A_1$ &             & $ +0.25~A_1(030) +0.24~A_1(040) +0.13~A_1(120) -0.13~A_1(050)$ \\
$  2143$ & $A_1$ &             & \\ 
$  2275$ & $A_1$ &  $A_1(120)$ & $ +0.86~A_1(120) +0.12~A_1(040)$ \\
$  2363$ & $A_1$ &             & $ -0.62~A_1(040) +0.40~A_1(120) -0.17~A_1(030) +0.11~A_1(050)$ \\
$  2436$ & $A_1$ &             & $ +0.47~A_1(040) +0.12~A_1(120)$ \\
$  2453$ & $B_2$ &  $B_2(020)$ & $ +0.53~B_2(020) +0.24~B_2(010) -0.17~B_2(030)$ \\
$  2576$ & $B_2$ &             & $ +0.27~B_2(020)$ \\
$  2735$ & $B_2$ &             & $ +0.60~B_2(020)$ \\
$  2743$ & $A_1$ &             & \\ 
$  2862$ & $A_1$ &             & \\ 
$  2903$ & $A_1$ &  $A_1(130)$ & $ -0.81~A_1(130) +0.12~A_1(050) -0.10~A_1(120)$ \\
$  2969$ & $A_1$ &             & \\ 
$  2977$ & $B_2$ &             & $ -0.24~B_2(110) -0.16~B_2(020)$ \\
$  3006$ & $B_2$ &             & \\ 
$  3045$ & $A_1$ &  $A_1(050)$ & $ +0.76~A_1(050) +0.13~A_1(130)$ \\
        \hline
    \end{tabular}
\end{table}

@@

We first discuss the comparison of our simulations to the lower-resolution
experimental spectrum presented on Figure~\ref{fig:sim_vs_dyke}. The peak A is
known to be a hot band, and does not appear in our ($0$~K)
simulation.~\cite{KDC:O3:92} The simulation reproduces well the consecutive
increase in the intensities of peaks B, C, D, and E. The spacing between these
peaks is also well reproduced. The drop in the intensity at the peak F is
captured by the simulation. Starting from the peak G, the simulation shows
discrepancy with the experiment. A sudden drop in the intensity past the peak
H is not observed in the simulation. A likely source of this mismatch is
discussed later.
